---
title: "Merge datasets: StreamPulse, Liu, and more"
output:
  pdf_document: default
  html_document: default
date: "2024-11-06"
---

## Intro info
Matt Nichols wrote the original code to merge the StreamPulse GPP dataset with the Liu et al. dataset. He had *already* used the Liu et al. geotiffs (and ArcGIS?) to merge some data by the StreamPulse lat longs, to create the here(data/aquatic/streams/streampulse_liu_merge.csv') file. This script merges the streampulse_liu_merge.csv file with two other files from Liu et al.: co2f.csv and k.csv, both in the here(data/Liu_data_DL_v20210902) folder. Apparently all the Liu files have COMID as a shared reference category. 

Christa Torrens amended the script to add this information, packages, and to make the read-data paths match the current folder structure, using the 'here' package.  We're still missing some details on the methods for creating streampulse_liu_merge (uploaded in this script), and the files that were used for that merge.  Asking Matt for the files and info 5/2/25 (he's defending soon, so TBD when he'll be able to share)

The correct/ complete stream datasets to use are found in the here('data/aquatic/streams/AutotrophyProjectFiles) folder. In that folder, the streampulse dataset is 'across_sites_model_data_annual.csv'. The mean monthly K600 values for each site-year-month are in 'StreamPulse_monthly_K.csv'. More information about these and other files in the folder is found in the AutotrophyProjectFiles_README.md file. 

NB: to run this code, you will need a couple of Liu et al. files, "co2f.csv" and "k.csv", that are too large to track with GitHub; instead, download them from the group's Google Drive folder/ Liu_data_DL and add them to your gitignore folder. You can also download them  yourself from DataDryad: https://datadryad.org/dataset/doi:10.5061/dryad.d7wm37pz9



```{r}

library(tidyverse)
library(here)

```


```{r}
#upload datasets

setwd()

streamPulse <- read.csv(here("data/aquatic/stream_pulse/20210902_streampulse_synthesis_statset.csv"), header = TRUE, sep = ",")

liu_co2 <- read.csv(here("data/terrestrial/co2f.csv"), header = TRUE, sep = ",")

#merge datasets based on COMID column

stream_merge <- merge(streamPulse, liu_co2, by = "COMID")

```

```{r}
#upload and add the Lui et al and streamPulse dataset

streampulse_liu_co2 <- read.csv(here("data/aquatic/streams/streampulse_liu_merge.csv"), header = TRUE, sep = ",")

k <- read.csv(here("k.csv"), header = TRUE, sep = ",")  ## PATH? 

```

```{r}
streamP_Liu_CO2 <- merge(streampulse_liu_co2, k, by = "COMID")

write_csv(streamP_Liu_CO2, file=here("data/aquatic/streams/MERGEDstreamPulse_Liu.csv"))
```


