---
title: "Merge datasets: StreamPulse, Liu, and more"
output:
  pdf_document: default
  html_document: default
date: "2024-11-06"
---

## Intro info

Matt Nichols started the code to merge the StreamPulse GPP dataset with the Liu et al. dataset. He had *already* used the Liu et al. geotiffs (and ArcGIS?) to manually merge the Liu et al. identifier COMID column with StreamPulse data, using the StreamPulse lat longs. The output product is the here(data/aquatic/streams/streampulse_liu_merge.csv') file. 

Christa Torrens completed and streamlined the script, pointing to the correct files and ensuring that the StreamPulse aggregated monthly K600 values were added to the merged dataframe. Other amendments inlude adding this information, R packages, and changing the read-data paths match the current folder structure by using the 'here' package. 

This script merges the streampulse_liu_merge.csv file with two other files from Liu et al.: co2f.csv and k.csv, both in the here(data/Liu_data_DL_v20210902) folder**. Apparently all the Liu files have COMID as a shared reference category. 

We're still missing some details on the methods for creating streampulse_liu_merge (uploaded in this script), and the files that were used for that merge.  Asking Matt for the files and info 5/2/25 (he's defending soon, so TBD when he'll be able to share)

The correct/ complete stream datasets to use are found in the here('data/aquatic/streams/AutotrophyProjectFiles) folder. In that folder, the streampulse dataset is 'across_sites_model_data_annual.csv'. The mean monthly K600 values for each site-year-month are in 'StreamPulse_monthly_K.csv'. More information about these and other files in the folder is found in the AutotrophyProjectFiles_README.md file. 

**NB: to run this code, you will need two  Liu et al. files, "co2f.csv" and "k.csv", that are too large to track with GitHub; instead, download them from the group's Google Drive folder/ Liu_data_DL and add them to your .gitignore file. You can also download them  yourself from DataDryad: https://datadryad.org/dataset/doi:10.5061/dryad.d7wm37pz9



```{r}

library(tidyverse)
library(here)

```


```{r}
#upload datasets

streamPulse_summary_df <- read.csv(here("data/aquatic/stream_pulse/20210902_streampulse_synthesis_statset.csv"), header = TRUE, sep = ",")

liu_co2 <- read.csv(here("data/terrestrial/co2f.csv"), header = TRUE, sep = ",")

#merge datasets based on COMID column

stream_merge <- merge(streamPulse, liu_co2, by = "COMID")

```

```{r}
#upload and add the Lui et al and streamPulse dataset

streampulse_liu_co2 <- read.csv(here("data/aquatic/streams/streampulse_liu_merge.csv"), header = TRUE, sep = ",")

k <- read.csv(here("k.csv"), header = TRUE, sep = ",")  ## PATH? 

```

```{r}
streamP_Liu_CO2 <- merge(streampulse_liu_co2, k, by = "COMID")

write_csv(streamP_Liu_CO2, file=here("data/aquatic/streams/MERGEDstreamPulse_Liu.csv"))
```

Checking sitenames for consistency between liu and streampulse K
```{r}




```

